\sys: Light-touch Data-Parallel Shell Processing

### 2020-05-11 pldi/draft.tex
Pass on 4.1

The core of \sys is a dataflow graph model, inspired by the ones used
in popular distributed stream processing systems~\sx{related} \kk{Is
  this reference here necessary?}. The dataflow model can express a
large subset of Shell programs and inherently supports the data
parallelism found in shell pipelines by exposing it as task
parallelism. More preciely, we have developed a set of
parallelization-exposing optimizations represented as
semantics-preserving graph transformations. These transformations can
then be applied in an iterative way, optimizing the program to utilize
available computational resources.

@@

\subsection{Graph Components}
\label{graph-components}

As mentioned earlier~\sx{bg:pipelines}, the two main abstractions of
the shell are (i) files (or pipes, \emph{viz.} correspondence between
files and pipes in \S\ref{bg:pipelines}) that contain data, and (ii)
commands that communicate through these files. Based on that, \sys's
graph model defines nodes to represent commands and edges to represent
files.

\paragraph{Edges---Files}

Before describing the graph edges, we introduce basic notation that
will be useful later on.  For a set $D$, we write $D\kstar$ to denote
the set of all finite words over $D$. For words $x, y \in D\kstar$, we
write $x \cdot y$ or $xy$ to denote their concatenation. We write
$\eps$ for the empty word. We say that $x$ is a \emph{prefix} of $y$,
and we write $x \leq y$, if there is a word $z$ such that $y = xz$. It
is easy to see that $\leq$ is a partial order, often called the
\emph{prefix order}.

Edges in the dataflow graph represent files, the basic data
abstraction of the shell. They are used as communication channels
between nodes in the graph, and as the input or output of the whole
graph. Edges are represented as possibly unbounded streams of type
$D\kstar$. They could either refer to a specific named file in the file
system, or just be FIFO pipes used for interprocess communication.

@@

%% \kk{I think we should delete this since the node and not the edge
%%   chooses how to write to its output files.}

%% In order to expose data parallelism from the output of a command,
%% files that do not point to a resource can have
%% \begin{wrapfigure}[3]{r}{0.20\columnwidth}
%% \vspace{-7pt}
%%   \includegraphics[width=0.20\columnwidth]{\detokenize{./figs/dish_g2.pdf}}
%% \vspace{-20pt}
%% \end{wrapfigure}
%% an upper bound on the lines that they can transfer. This helps utilize available
%% parallelism after encountering a command that cannot be parallelized
%% (\eg \ttt{sha1sum}). The schematic on the right illustrates this scenario.

\tr{On the other hand, some forms of data parallelism can be exposed
  when knowing the size of the input files. As mentioned in \ref{}
  some pure commands (such as cat -n) only need line information to
  become stateless, and knowing the size of a file could allow the
  system to split it in different chunks that can be processed
  independently. To account for that, edges that refer to an input
  resource contain the number of lines of the file that they refer
  to.}

Finally, edges that do not refer to a resource and do not start
from a node in the graph represent the graph input;
edges that do not refer to a resource and do not point to a node in the graph are its outputs.

\paragraph{Nodes---Commands}

A node $f$ of the graph represents a function from a (possibly empty)
list of input streams to a list of output streams $f : [D\kstar] \to
[D\kstar]$, where $D$ represents the basic data type of a line of
characters. This representation captures the majority of shell
commands. We require that the function $f$ is monotone w.r.t.\ a
lifting of the prefix order for a sequence of inputs \kk{Formalize
this lifting if essential for later development}. This captures the
  idea that a node cannot retract output after it has been emitted.
%More formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there
%exists $y'$ such that $f(x.x') = y.y'$, where $.$ is standard
%concatenation and will be omitted when obvious.
An example of a node with one input and one output stream is the
following command:

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 grep ``foo''
\end{lstlisting}

\noindent
This command takes a stream of lines from its standard input and
returns a stream of lines on its standard output.

An important observation is that even though commands might read their
input from several input streams, the order in which they read is not
always arbitrary. An illustrative example is the command \ttt{cat},
which consumes its input one stream at a time:

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 cat x - y
\end{lstlisting}

\noindent
In the example above, \ttt{cat} reads file \ttt{x} until it encounters
the EOF character, then reads from \ttt{stdin} until in encounters
EOF, and finally reads from \ttt{y}. Other commands that read their
inputs one file at a time include \ttt{grep} and \ttt{tr}. Knowing the
order in which a node consumes its input is necessary to enable
certain optimizations on the graph.

@@

\kk{I need to mention that there are nodes that we know more
  properties of. For example cat, tee, zip, etc... and these are
  handled specially because of their very important and special
  properties.}

@@

Having a dataflow graph representation allows us to define
parallelization optimizations as graph transformations. Before
describing the different types of optimizations, we formalize the
intuition about stateless and pure commands that was described
earlier~\sx{cmd}. We focus on parallelizing commands that produce one
output and consume their inputs one at a time, with the possible
exception of static input files that are used for configuration. More
formally, we parallelize commands that can be represented with
functions of type $f : D\kstar \times [D\kstar] \to D\kstar$, where
the first argument represents the concatenation of the inputs that are
consumed one at a time and the second argument represents the static
inputs. When translating a command that consumes its input streams one
at a time, a \ttt{cat} node is introduced before it to concatenate the
streams and its output is transferred to the command input. \kk{Maybe
  have an argument here that we show in the evaluation that handling
  these commands is adequate for a big variety of data processing
  scripts (and most stateless and pure commands), from several
  microbenchmarks to some large-realistic macrobenchmarks.}.
%% \nv{TODOs
%%   (i) highlight earlier the antithesis between the formalisms in this section and the informalities of the previous (me)
%% }

>>>>

The core of \sys is a dataflow graph model that is expressive enough
to support a large subset of shell programs. A fundamental difference
with other dataflow graph models is that in \sys's model contains
information about the order in which a node reads its inputs. This
enables a set of graph transformations that expose data parallelism as
task parallelism. These transformations can then be applied in an
iterative way, optimizing the program to utilize available
computational resources.

%% The core of \sys is a dataflow graph model, inspired by the ones used
%% in popular distributed stream processing systems~\sx{related} \kk{Is
%%   this reference here necessary?}. The dataflow model can express a
%% large subset of Shell programs and inherently supports the data
%% parallelism found in shell pipelines by exposing it as task
%% parallelism. More preciely, we have developed a set of
%% parallelization-exposing optimizations represented as
%% semantics-preserving graph transformations. These transformations can
%% then be applied in an iterative way, optimizing the program to utilize
%% available computational resources.

@@

\subsection{Definitions}
\label{graph-components}

The two main shell abstractions are (i) files (or pipes, \emph{viz.}
correspondence between files and pipes in \S\ref{bg:pipelines}) that
contain data, and (ii) commands that communicate through these
files. Based on that, in \sys's graph model nodes represent commands
and edges represent files.

We first introduce basic notation that will be useful later on.  For a
set $D$, we write $D\kstar$ to denote the set of all finite words over
$D$. For words $x, y \in D\kstar$, we write $x \cdot y$ or $xy$ to
denote their concatenation. We write $\eps$ for the empty word. We say
that $x$ is a \emph{prefix} of $y$, and we write $x \leq y$, if there
is a word $z$ such that $y = xz$. The $\leq$ order is reflexive,
antisymmetric, and transitive (i.e. is a partial order) and is often
called the \emph{prefix order}.

\paragraph{Edges -- Files}

Edges in the dataflow graph represent files, the basic data
abstraction of the shell. They are used as communication channels
between nodes in the graph, and as the input or output of the whole
graph. Edges are represented as possibly unbounded streams of type
$D\kstar$. Edges can either refer to named files or FIFO pipes used
for interprocess communication. Edges that do not start from a node in
the graph represent the graph input; edges that do not point to a node
in the graph are its outputs.

@@

%% \tr{On the other hand, some forms of data parallelism can be exposed
%%   when knowing the size of the input files. As mentioned in \ref{}
%%   some pure commands (such as cat -n) only need line information to
%%   become stateless, and knowing the size of a file could allow the
%%   system to split it in different chunks that can be processed
%%   independently. To account for that, edges that refer to an input
%%   resource contain the number of lines of the file that they refer
%%   to.}

\paragraph{Nodes -- Commands}

A node $f$ of the graph represents a function from a (possibly empty)
list of input streams to a list of output streams $f : [D\kstar] \to
[D\kstar]$, where $D$ represents the basic data type of a line of
characters. This representation captures the majority of shell
commands. We require that the function $f$ is monotone w.r.t.\ a
lifting of the prefix order for a sequence of inputs \kk{Formalize
  this lifting if essential for later development}.
%% More formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there
%% exists $y'$ such that $f(x.x') = y.y'$, where $.$ is standard
%% concatenation and will be omitted when obvious.
This captures the idea that a node cannot retract output that it has
already produced.

\paragraph{\todo{One-input-output} Commands}

\kk{Find a better name.} \todo{One-input-output} commands \todo{are
  special commands} that consume their inputs one at a time, with the
possible exception of static input files, and produce one output
file. These commands can be represented as functions of type $f :
D\kstar \times [D\kstar] \to D\kstar$, where the first argument
represents the concatenation of the inputs that are consumed one at a
time and the second argument represents the static inputs. The
quintessential example of \todo{one-input-output} commands is
\ttt{cat}, that consumes its inputs one at a time, producing their
concatenation as output.
%% A \todo{large subset} of shell commands fall in this category

\kk{Maybe give an example of a script and its dataflow graph.}

%% An example of a node with one input and one output stream is the
%% following command:

%% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
%%  grep ``foo''
%% \end{lstlisting}

%% \noindent
%% This command takes a stream of lines from its standard input and
%% returns a stream of lines on its standard output.

%% An important observation is that even though commands might read their
%% input from several input streams, the order in which they read is not
%% always arbitrary. An illustrative example is the command \ttt{cat},
%% which consumes its input one stream at a time:

%% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
%%  cat x - y
%% \end{lstlisting}

%% \noindent
%% In the example above, \ttt{cat} reads file \ttt{x} until it encounters
%% the EOF character, then reads from \ttt{stdin} until in encounters
%% EOF, and finally reads from \ttt{y}. Other commands that read their
%% inputs one file at a time include \ttt{grep} and \ttt{tr}. Knowing the
%% order in which a node consumes its input is necessary to enable
%% certain optimizations on the graph.

@@

%% \kk{I need to mention that there are nodes that we know more
%%   properties of. For example cat, tee, zip, etc... and these are
%%   handled specially because of their very important and special
%%   properties.}

@@

Having a dataflow graph representation allows us to define
parallelization optimizations as graph transformations. Before
describing the different types of optimizations, we formalize the
intuition about stateless and pure commands that was described
earlier~\sx{cmd}.
%% \nv{TODOs
%%   (i) highlight earlier the antithesis between the formalisms in this section and the informalities of the previous (me)
%% }

\kk{Maybe
  have an argument here that we show in the evaluation that handling
  these commands is adequate for a big variety of data processing
  scripts (and most stateless and pure commands), from several
  microbenchmarks to some large-realistic macrobenchmarks.}.
